---
title: "Homework 3"
author: "Shreya Rao  sr3843"
date: "5/27/2021"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#Loading libraries
library(plyr)

#Loding the file
debt <- read.csv("debt.csv", as.is = TRUE)
dim(debt)
head(debt)
#ratio: debt to GDP ratio
```


1. Calculate the average GDP growth rate for each country (averaging over years). This
is a classic split/apply/combine problem, and you will use daply() to solve it.

```{r}
#a)
mean.growth <- function(df) {
  return(signif(mean(df$growth, na.rm = TRUE), 3))
}

#b)
avg_growth <- daply(debt, .(Country), mean.growth)
avg_growth["Australia"]
avg_growth["Netherlands"]
```


2. Using the same instructions as problem 1, calculate the average GDP growth rate for
each year (now averaging over countries). (The average growth rates for 1972 and
1989 should be 5.63 and 3.19, respectively. Print these values in your output.) Make
a plot of the growth rates (y-axis) versus the year (x-axis). Make sure the axes are
labeled appropriately.

```{r}
year_means <- daply(debt, .(Year), mean.growth)
year_means[c("1972", "1989")]

plot(x=names(year_means), y=year_means, xlab = "Year", ylab="Growth Rate")
```


3. The function cor(x,y) calculates the correlation coefficient between two vectors x
and y.

a. Calculate the correlation coefficient between GDP growth and the debt ratio over
the whole data set (all countries, all years). Your answer should be −0.1995.

```{r}
signif(cor(debt$growth, debt$ratio), 4)
```

b. Compute the correlation coefficient separately for each country, and plot a histogram of these coefficients (with 10 breaks). The mean of these correlations
should be −0.1778. Do not use a loop. (Hint: consider writing a function and
then making it an argument to daply()).

```{r}
cor_func <- function(df) {
  return(signif(cor(df$growth, df$ratio), 3))
}

country_cor <- daply(debt, .(Country), cor_func)
#country_cor
signif(mean(country_cor), 4)

hist(country_cor, breaks = 10, xlab="Country Correlation Coefficient")
```

c. Calculate the correlation coefficient separately for each year, and plot a histogram
of these coefficients. The mean of these correlations should be −0.1906.

```{r}
year_cor <- daply(debt, .(Year), cor_func)
signif(mean(year_cor), 3)

hist(year_cor, breaks = 10, xlab="Year Correlation Coefficient")
```

d. Are there any countries or years where the correlation goes against the general
trend?

```{r}
country_cor
country_cor[country_cor > 0.2 | country_cor < -0.6]
```
Norway has an unsually high positive correlation of 0.563 while the rest have correlation close to 0 or negative correlation.
Japan and Italy have high negative correlations (below -0.65). 

```{r}
#plot(names(year_cor), year_cor)
year_cor[year_cor > quantile(year_cor, 0.975) | year_cor < quantile(year_cor, 0.025)]

```
The years 1946 and 1975 have high correlations that are negatively correlated (below -0.6).
1978 has an evidently high postive correlation of 0.431.


4. Fit a linear model of overall growth on the debt ratio, using lm(). Report the intercept
and slope. Make a scatter-plot of overall GDP growth (vertical) against the overall
debt ratio (horizontal). Add a line to your scatterplot showing the fitted regression
line

```{r}
lm_growth_debt <- lm(growth~ratio, data=debt)
plot(x=debt$ratio, y=debt$growth, xlab="Debt Ratio", ylab="GDP Growth")
abline(lm_growth_debt, col="blue")
```


5. There should be four countries with a correlation smaller than -0.5. Separately, plot
GDP growth versus debt ratio from each of these four countries and put the country names in the titles. This should be four plots. Call par(mfrow=c(2,2)) before
plotting so all four plots will appear in the same figure.

```{r}
countries_4 <- country_cor[country_cor < -0.5]
names_4 <- names(countries_4)
par(mfrow=c(2,2))

for(i in names_4) {
  plot(y=debt$growth[debt$Country == i], x=debt$ratio[debt$Country == i], xlab="Debt Ratio", ylab="GDP Growth")
  title(main = i)
}
```


6. Some economists claim that high levels of government debt cause slower growth. Other
economists claim that low economic growth leads to higher levels of government debt.
The data file, as given, lets us relate this year’s debt to this year’s growth rate; to
check these claims, we need to relate current debt to future growth.

a. Create a new data frame which just contains the rows of debt for France, but
contains all those rows. It should have 54 rows and 4 columns (print the dimensions of your data frame). Note that some years are missing from the middle of
this data set.

```{r}
france <- debt[debt$Country == "France",]
dim(france)
```

b. Create a new column in your data frame for France, next.growth, which gives
next year’s growth if the next year is in the data frame, or NA if the next year is missing. (next.growth for 1971 should be (rounded) 5.886, but for 1972 it
should be NA. Print these two values.)
```{r}

for(i in 1:(nrow(france))) {
  if ((france$Year[i]+1 == france$Year[i+1]) & (i != nrow(france))) { 
    france$next.growth[i] <- signif(france$growth[i+1], 4)
  }
  else { 
    france$next.growth[i] <- NA 
  }
}

france[france$Year == 1971 | france$Year == 1972,]
```


7. Add a next.growth column, as in the previous question, to the whole of the debt
data frame. Make sure that you do not accidentally put the first growth value for one
country as the next.growth value for another. (The next.growth for France in 2009
should be NA, not 9.167. Print this value.) Hints: Write a function to encapsulate
what you did in the previous question, and apply it using ddply().

```{r}
next_growth <- function(df) {
  
  temp_growth = NULL
  for(i in 1:(nrow(df))) {
    if ((df$Year[i]+1) == df$Year[i+1] & (i != nrow(df))) { 
      df$next.growth[i] <- round(df$growth[i+1], 3)
    }
    else { 
      df$next.growth[i] <- NA 
    }
  }
  return(df)
}

# t_lists <- split(debt, debt$Country)
# sapply(t_lists, next_growth)

debt <- ddply(debt, .(Country), next_growth)

debt[debt$Country == "France" & debt$Year == 2009,]
```


8. Make a scatter-plot of next year’s GDP growth against this year’s debt ratio. Linearly
regress next year’s growth rate on the current year’s debt ratio, and add the line to the
plot. Report the intercept and slope to reasonable precision. How do they compare
to the regression of the current year’s growth on the current year’s debt ratio?

```{r}
plot(x=debt$ratio, y=debt$next.growth, xlab="Current Debt Ratio", ylab="Next Year's Growth Rate")

reg_line <- lm(next.growth~ratio, data = debt)
abline(reg_line, col="purple")
#Next year's coefficients
signif(coefficients(reg_line), 3)
#summary(reg_line)

#This year's coefficients
signif(coefficients(lm_growth_debt), 3)
```
Both of them have similar negative slopes, with the current year's regression slope very slightly less that next year's. They also have similar intercepts around 4. 
Overall, they have similar regression lines. 


9. Make a scatter-plot of next year’s GDP growth against the current year’s GDP growth.
Linearly regress next year’s growth on this year’s growth, and add the line to the plot.
Report the coefficients. Can you tell, from comparing these two simple regressions
(from the current question, and the previous), whether current growth or current debt
is a better predictor of future growth?

```{r}
plot(x=debt$growth, y=debt$next.growth, xlab="Current Growth", ylab="Next Year Growth")
growth_reg <- lm(next.growth~growth, data = debt)
abline(growth_reg, col="red")

signif(coefficients(growth_reg), 3)
summary(growth_reg)$r.squared

#Next year vs debt summary
summary(reg_line)$r.squared
```
There seems to be a more linear relationship between next year's growth and current year's growth. This is also indicated by the slopes. This regression line also seems to be a better fit (as indicated by a higher R squared value) than that of next year's growth against current debt. 
Thus, current growth seems to be a better predictor of future growth. 
